### 2012-03-23 docs/notes.tex
Outline REM DP c++ and add tests.
Edit notes for final report.

\abstract{
Stochastic blockmodels assume latent groups of individuals have similar tendencies in their  group-wise interactions.  Many find these methods useful for exploratory analysis of static networks. On the other hand, continuous-time methods such as the relational event framework (Butts 2008) model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.  We propose to combine these two approaches where each pair of blocks has a set of parameters guiding the event dynamics for dyads between those blocks.  The method is illustrated using dyadic interaction data (e.g. email, Twitter direct messages) among hundreds of actors and thousands of events.  We show parameter estimates from the model can reveal heterogeneity in the dynamics among groups of individuals, providing a better fit to the data than using a single relational event model.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.

@@

Network data, however, is often collected as a sequence of events occurring over time.   Recently several works have adapted models from event history analysis \cite{} to provide continuous time models of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.  These models allow one to specify how the process depends on the previous history of events.  In this way one may incorporate theories about the underlying processes and make predictions about future data conditioned on the past.

The above continuous time network models have two major drawbacks: 1) the entire network is assumed to have identical dynamics, and 2) the occurrence of any event can affect the rate at which any other event occurs.  Both aspects may be unrealistic.  Consider email communication in a small organization comprised of several teams.  Each team may have a different pattern for collaborating via email, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.  The behavior of the model is illustrated with data simulated from the model.  Using  data sets of dyadic communication among people, we demonstrate the ability to improve upon simple baselines for predicting the next event.  Finally we show that the parameter estimates provide an interpretable structure to the event dynamics among blocks of individuals, sometimes identifying particular \emph{roles} of communication that exist within the network.

@@

One can extend the above to marked point processes where each event contains additional information.  In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ of possible events include all allowed dyads.  Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t)$ that is piecewise constant.  We can then write the likelihood as

\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m}) \lambda_{ij}(t_m | z_i,z_j) \}
\end{align}
\noindent where event $m$ is $(i_m,j_m)$, $\tau_{i,j,m}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads who have a changepoint in their intensity when $(i_m,j_m)$ occurs. \footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
}
\caption{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4. [TODO? Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.]}
\label{fig:example}
\end{figure}

@@

We hope to learn about the dynamics between subsets of nodes while allowing for variety among disjoint subsets.   To facilitate this, we assume each node belongs to a cluster $z_i \sim \mbox{Categorical}(\alpha)$ where $\sum_k \alpha_k = 1$ and  model the intensity functions via a log linear model
\begin{align}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z}).
\end{align}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}_{ij}(t|\cdot)$.   Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.

@@

We construct $\mathbf{s}(t,i,j)$ using a similar set of statistics as used in \cite{Butts2008,Vu2011}. Let $v_{ijm} \in [0,M]$ be the index of the last event where $(i,j)$ occurred (n.b. $t_{v_{ijm}} = \tau_{ijm}$). If $(i,j)$ has never occurred, we set $v_{ijm}=0$.  
\begin{enumerate}
\item Intercept: $s_{0}(t,i,j) = 1$
\item Dyad count: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i) / m$
\item Sender out-degree: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i) / m$
\item Sender in-degree: $s_{2}(t,i,j) = \sum_{m:t_m<t} I(j_m=i) / m$
\item Receiver out-degree: $s_{3}(t,i,j) = \sum_{m:t_m<t} I(i_m=j)/m$
\item Receiver in-degree: $s_{4}(t,i,j) = \sum_{m:t_m<t} I(j_m=j)/m$
\item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
%\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
% \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
% \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
% \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
\end{enumerate}

We include degree effects to model possible rich-get-richer phenomenon.  For example, a particular node sending often may indicate they will continue to be a sender in the future.  We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Develop some theory about these sorts of statistics.}

@@

Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.  Here, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

The above framework generalizes several important special cases.  The first is one where events within each block share the rate at which they occur.  This can be implemented by only having an intercept statistic, $s_0(t,i,j) = 1$.  As these intensities do not change under this specification, the likelihood simplifies to 
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
where $ \lambda_{ij} = \exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  Analogous to the stochastic block model for static networks, this model posits each dyad is a homogeneous Poisson process and all dyads within block $(z_i,z_j)$ have the intensity $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.

@@

We use Markov chain Monte Carlo to get samples from the posterior distribution of our parameters.

\begin{enumerate}
\item For $i \in [1,N]$: Sample $z_i$ via Gibbs sampling: 
$z_i | A, z_{-i} \propto p(A|z,\beta) p(z_i)$.  (See Appendix.)
\item For each pair $(k,k')$: Sample $\beta_{k,k'}$ via MH or HMC.
\end{enumerate}

@@

Slice sampling with $m=20$.

 To perform Gibbs sampling for a given cluster assignment $z_i$, we need to compute the likelihood for each possible assignment of $z_i$.   By limiting the changepoints to times when either node is involved, computing the likelihood is $O(T \cdot P \cdot 4N)$ where 4 comes from considering all dyads where either $i$ or $j$ is involved.  We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m,i,j$.  Note that 
% This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.  Computing the likelihood is then $O(T \cdot P \cdot 4N)$ (though this computation is split into $K^2$ pieces) and must computed $N \cdot K$ times per Gibbs iteration.  The most straightforward implementation has space complexity $O(N^2)$ since we must keep track of $\tau_{i,j,m}$ for each $(i,j)$ at a given $m$.

@@

\section{Simulation}
\begin{figure}
\center
\includegraphics[width=2in]{../figs/synthetic/mat.pdf}
\includegraphics[width=3in]{../figs/synthetic/counts.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Right: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.}
\label{fig:syncounts}
\end{figure}

We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to continue speaking, and 3) interactions between groups are more likely to be repeated.  The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{5}(t,i,j), s_{9}(t,i,j), s_{11}(t,i,j)]$.  For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (2,3,0,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (1,0,0,1)$, and $\boldsymbol{\beta}_{2,2} = (2,0,2,0)$.  See Appendix for details on simulating from the model.  In Figure \ref{fig:syncounts} we illustrate the simulated data and show the loglikelihood during the first iterations of MCMC [after being initialized with $\beta \sim N(\beta_{true},.5^2)$].  Note that we recover the same cluster assignments and the full model converges to the true loglikelihood.

TODO: Show parameter bias as a function of iteration; perhaps also as a function of the number of events in the dataset.

@@

@@

\begin{itemize}
%\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.  We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.  Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
\item The Irvine data set
\end{itemize}

@@

\begin{figure}
\center
%\includegraphics[width=3in]{../figs/synthetic/logposterior.pdf}
\includegraphics[width=3in]{../figs/synthetic/test-recall.pdf}
\caption{Results from fitting models to synthetic data. Left: Log posterior of 1000 simulated events vs. iteration during MCMC.  Right: Recall on 3000 test events.}
\label{fig:synresults}
\end{figure}
We evaluate the predictive ability of the learned models by comparing models based on the loglikelihood of held-out data and recall on held out data.  Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.

@@

Our method jointly models both which dyads occur and when they occur.  To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j} = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.

>>>>

\abstract{
Observing network data as a stream of events provides an opportunity to understand differences in how nodes behave given the past.  Recent methods for continuous-time network data model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.  Methods for static networks, such as stochastic blockmodels, often account for actor-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.  We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  The method is illustrated using dyadic interaction data -- e.g. email, Twitter direct messages.  We show parameter estimates from the model reveal heterogeneity in the dynamics among groups of individuals.  The fitted models have better predictive accuracy than both baselines and relational event models without the latent structure.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.

@@

Network data, however, is often collected as a sequence of events occurring over time.   Recently work leverages  continuous time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.  These models allow one to specify how the process depends on the previous history of events.  In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.

The above continuous time network models assume all nodes in the network behave according to identical dynamics.  In many cases this is unrealistic.  For example, in a small organization comprised of several teams, each team may exhibit different patterns for collaborating via email --  individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.  %, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  The behavior of the model is illustrated with simulated data.  We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  Finally we show that the parameter estimates provide an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication that exist.

@@

One can extend the above to marked point processes where each event contains additional information.  In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.  We assume each dyad $(i,j) \in \mathcal{R}$ has an intensity function $\lambda_{ij}(t)$ that is piecewise constant.  We can then write the likelihood as  
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | z_i,z_j) \}
\label{eqn:llk}
\end{align}
\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{i,j,m}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs. \footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
}
\caption[]{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.\footnotemark }
\label{fig:example}
\end{figure}

\footnotetext{TODO? Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}

@@

We hope to learn about the dynamics between subsets of nodes while allowing for variety among disjoint subsets.   To facilitate this, we assume node $i$ is assigned a latent cluster $z_i$ and the intensity functions via a log linear model
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) &= \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t) &
z_i &\sim \mbox{CRP}(\alpha) &
\beta_{k,l,p} &\sim \mbox{N}(\mu_p,\sigma_p^2)
\end{align*}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}(t,i,j|\cdot)$.   Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.  To allow for the blocks to share information by placing a hierarchical prior on each $\boldsymbol{\beta}_{k.l}$.

@@

% \begin{enumerate}
% \item Intercept: $s_{0}(t,i,j) = 1$
% \item Dyad count: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
% \item Sender out-degree: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
% \item Sender in-degree: $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$
% \item Receiver out-degree: $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$
% \item Receiver in-degree: $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$
% \item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% \item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% \item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% \item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% \item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% %\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
% % \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
% % \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
% % \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
% \end{enumerate}

\begin{table}[t]
\footnotesize
\center
\begin{tabular}{|l|l|}
\hline
Statistic & Formula \\
\hline
\hline
Intercept:& $s_{0}(t,i,j) = 1$\\
Dyad count:& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
Sender out-degree:& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
Sender in-degree:& $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
Receiver out-degree:& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
Receiver in-degree:& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
Reciprocity (AB-BA):& $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$\\
Turn-taking (AB-BY): &$s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}\ne i)$\\
Turn-usurping (AB-XA):& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}} \ne j,j_{v_{ijm}}=i)$\\
Turn-usurping (AB-XB): &$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}} \ne i,j_{v_{ijm}}=j)$\\
Turn-continuing (AB-AY):& $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}\ne j)$\\
\hline
\end{tabular}
\label{tab:stats}
\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
\end{table}

In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred (n.b. $t_{v_{mij}} = \tau_{mij}$). If $(i,j)$ has never occurred, we set $v_{mij}=0$.  Here we include degree effects to model possible rich-get-richer phenomenon.  For example, a particular node sending often may indicate they will continue to be a sender in the future.  We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Explore the theory about these sorts of statistics.  Aalen has a few suggestions.} Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}

@@

Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.  In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

The above framework generalizes several important special cases. For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.  Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  As these intensities do not change under this specification, the likelihood simplifies to 
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$

Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood becomes a conditional logit
$$\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.$$
The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.

\subsection{Related work}

Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant and allow for a nonlinear relationship between statistics and intensity functions.

(TODO: More here.  For example: \cite{Tillman2011} explores the idea of having nonpoisson processes at each edge and solves the master equation for the dynamical system.)

@@

We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.

\subsection*{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }

We Gibbs sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_i | z_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto p(z_i | z_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}).% & z_i &= \mbox{new} = \alpha
\end{align*}
Letting $\mathcal{U}_r = \{a: r \in \{i_a,j_a\}, a \in [1,M]\}$ be the set of events where node $r$ is involved, the likelihood term above can be written as
$$p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) = \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}$$

\noindent Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$. 
% \begin{align*}
% p(z_i = k | z_{-i},\alpha) &= n_k & p(z_i = \mbox{new} |  z_{-i},\alpha) &= \alpha
% \end{align*}

 By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot 4N)$.  The factor of 4 is due to the number of dyads involving a single node.  We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m,i,j$.

\subsection*{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }

For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
\begin{align*}
p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \mu, \sigma) &\propto p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\  
p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)
\end{align*}
the likelihood term above can be written as
$$p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) = \prod_{a \in \mathcal{V}_{k,l}} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}$$
\noindent where $\mathcal{V}_{k,l} = \{a: z_{i_a} \in \{k,l\} \mbox{or} z_{j_a} \in \{k,l\}, a \in [1,M]\}$ is the set of events $a$ where one of the nodes is either in block $k$ or $l$.  \footnote{TODO: If we end up using HMC, mention step size and step length used.}
The above distribution is sampled using slice sampling.\footnote{We may also use HMC for this.}

\subsection*{Split-merge algorithm}

Motivation: simply Gibbs sampling does not seem to mix well.  \footnote{This section is planned future work.}

In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.  For these chains we do not propose split moves when the number of non-empty clusters is already $K$.

@@

% This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.  Computing the likelihood is then $O(T \cdot P \cdot 4N)$ (though this computation is split into $K^2$ pieces) and must computed $N \cdot K$ times per Gibbs iteration.  The most straightforward implementation has space complexity $O(N^2)$ since we must keep track of $\tau_{i,j,m}$ for each $(i,j)$ at a given $m$.

@@

\begin{figure}
\center
\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
\includegraphics[width=2.25in]{../figs/synthetic/counts.pdf}
\includegraphics[width=2.5in]{../figs/synthetic/params-estimates.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.  Right: Parameters (in red) and posterior credible intervals (in black).}
\label{fig:syncounts}
\end{figure}

\section{Simulation}

We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to continue speaking, and 3) interactions between groups are more likely to be repeated.  The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{5}(t,i,j), s_{9}(t,i,j)]$.  For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,0,2)$.  This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.

In Figure \ref{fig:syncounts} we illustrate the simulated data.  While the dyad counts suggest a stochastic blockmodel, the center plot shows each block has empirical differences in their dynamics: intensities for reciprocal actions among actors in block 1 are $e^3$ times greater, intensities for turn-taking actions among actors in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.  Fitting the model with $K=2$ converges to the true loglikelihood (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.

@@

\small{
\input{../figs/results.tex}
}

@@

\begin{itemize}
%\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
\item Eckmann email:
\item Kiel email: Still need to hunt this down.
\item Enron email:
\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.  We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.  Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
\item MIT Reality Mining: 
\end{itemize}

@@

\begin{figure}[th]
\caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
\label{fig:recall}
\end{figure}

%\begin{figure}
%\center
%\includegraphics[width=3in]{../figs/synthetic/logposterior.pdf}
% \includegraphics[width=3in]{../figs/synthetic/test-recall.pdf}
% \caption{Results from fitting models to synthetic data. Left: Log posterior of 1000 simulated events vs. iteration during MCMC.  Right: Recall on 3000 test events.}
% \label{fig:synresults}
% \end{figure}
We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.  Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.  I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}

@@

Our method jointly models \emph{which} dyads occur and \emph{when} they occur.  To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j} = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
